\vspace{-2cm} Inference on the Effects of Observed Features \\ in Latent Space Models for Networks

### 2017-09-06 manuscript_rearranged.tex
Hanna's edits.

\noindent Due to the complex interdependence found in network data, political networks scholars draw upon an increasingly sophisticated toolkit for statistical inference. The latent space model (LSM) for network data combines a generalized linear model with a latent spatial embedding of the network. In numerous applications, researchers assume that the latent spatial embedding can control for unmeasured confounding structure that is related to the values of edges in the network. There has been little to no research that considers the LSM's performance in adjusting for unmeasured structure. We investigate the LSM's performance via a simulation study. In the presence of an unmeasured covariate that can be modeled using a latent space, estimation and inferential error remain high under even moderate confounding. However, the prediction error of the LSM when unmeasured network structure is present is substantially lower in most cases. We conclude that the LSM is most appropriate for exploration or prediction.%\footnote{This work was supported by National Science Foundation grants DGE-1144860, SES-1558661, SES-1619644, and CISE-1320219. Any opinions, findings, and conclusions or recommendations are those of the authors and do not necessarily reflect those of the sponsor.}

\end{abstract}
\thispagestyle{empty}
% \doublespacing
% Description of the possible challenges
\section{Introduction}

@@

Despite their growing popularity, few studies exist that investigate the performance of these models in adjusting for confounding network structure.  The approach to adjusting for dependencies in the two other models commonly used for network data---ERGM and SAOM---is quite similar to adjusting for confounding covariates in regression modeling. The researcher specifies a set of dependencies that (s)he hypothesizes to be important in the generative model for the network. These dependencies are then explicitly included in a model that simultaneously represents the effects of observed covariates \citep{cranmer2011pa}. The LSM takes a different approach, which involves the incorporation of latent variables to model network structure. The LSM has an advantage over ERGM and SAOM in that researchers need not develop a set of hypothesized dependencies in order to model network structure that is not reflected in observed covariates. However, this advantage hinges upon the capacity for the LSM to discover unmodeled structure that could otherwise be attributed to the observed covariates. In the current study, we focus on the LSM, examining its performance in reducing estimation and inferential errors regarding the effects of observed covariates via adjustment for confounding network structure.

\subsection{Central Problem}

Latent variable inference, generally conceived, presents the possibility of representing unmeasured data in statistical models. The LSM, introduced by \citet{hoff2002latent}, is used to estimate the effect of covariates in the presence of latent network structure. Here the distance function $|z_i - z_j|$ represents latent network structure as homophily with respect to latent variables.  The distance function is additively combined with a regression on observed dyadic covariates, $x_{ij}$, to form a linear predictor for tie prediction.  As with a GLM, a link function, $g^{-1}$, maps this linear predictor to the appropriate edge distribution, giving

@@

This ``Euclidean'' LSM is the original form proposed by \citet{hoff2002latent}, and, as far as we can tell, the most commonly used specification in the literature. However, the LSM has been extended in several important directions. The  The latent space has been represented as a $k$-dimensional Euclildean space and by latent factors with a multiplicative metric \citep{hoff2002latent, hoff2004modeling, hoff_2005_jasa, hoff2009multiplicative}. Additional structure has been introduced by adding random effects, which, for example, may involve sender or receiver specific effects for directed networks which capture differential activity rates amongst nodes \citep{hoff2003random}. Within this framework \cite{westveld2011mixed} models dynamic network data by treating the latent space as a stochastic process. \cite{handcock2007model} enable the LSM to model clustering that is not representable as homophily (i.e., stochastic equivalence) by combining the LSM with latent cluster models. \cite{hoff2008modeling} shows that the LSM and latent cluster models are special cases of an ``eigenmodel.'' That is, an eigendecomposition of a symmetric sociomatrix can be used to represent both latent space and cluster models, but not vice-versa. \citet{hoff2015dyadic} consolidates, extends, and implements several of these extensions in the \texttt{amen} package \citep{amen} for \R \citep{minhas2016inferential, hoff2015dyadic}.

Despite the power of the LSM in representing network structure that is not modeled by the observed covariates, and its performance in predictive ties predicting ties, we have reason to doubt the LSM's capacity to adjust for confounding network structure. We do not claim that the methodological literature in which the LSM was introduced or extended has presented the LSM as capable of adjusting for unmeasured confounding structure. However, as we document below, many researchers who have applied the LSM claim that it has the capacity to adjust for confounders, and use it for this purpose. There are two reasons that using the LSM may lead to increased error---one may result in Type 1 inferential error, and the other in Type 2 error. First, the latent configurations inferred may result in a representation of the network wherein a node's position in the latent space is spuriously correlated with the observed covariates. Inferring latent variables that are correlated with observed covariates would lead to reduced efficiency in estimating observed covariate effects, and result in a high Type 2 error rate.  Second, if the unobserved (i.e., latent) network structure is truly correlated with the observed covariates, the unobserved structure that can be correlated with the observed variable may be attributed to the observed variable. Under this condition the latent space parameters are used to model sources of variation that cannot be attributed to the observed covariate, and inferences regarding the observed covariates remain subject to omitted variable bias, which leads to an inflated Type 1 inferential error rate.

Recent work on the LSM has explicitly addressed the potential problem of Type 2 error. The AMEN framework for latent variable modeling of networks is explicitly designed to model structure in the residuals---structure that is leftover after accounting for the observed covariates. \citet[p. 43]{hoff2015dyadic} notes that the latent factors represent patterns in the network that, ``aren't explained by the known regressors.'' \citet[pp. 12--13]{minhas2016inferential} also describe how AMEN is designed such that the multiplicative effects, ``capture higher-order dependence patterns that are left over [in the stochastic linear predictor] after accounting for any known covariate information.''  The approach incorporated in AMEN is effective at avoiding correlation between the observed covariates and latent factors, which would result in Type 2 error. However, there have been no methodological innovations to avoid Type 1 error in latent variable modeling with networks. Below we review several studies in which researchers have relied on the latent space modeling framework to adjust for unmeasured confounding structure.  Through a simulation study we find that the LSM, as implemented in both the R \citep{R} packages \texttt{latentnet} \citep{latentnet} and \texttt{amen} \citep{amen}, performs poorly in terms of Type 1 error when the unmeasured network structure confounds the relationship between the observed covariate and the network.

\section{Applications of the Latent Space Model}

@@

% need to add marketing stuff and probably should look again for applications
% also need to read and fit in the newer ward student stuff

Although the LSM has often been used as an exploratory or predictive model, it has been applied in many cases for explanatory causal modeling---to reduce estimation and/or inferential error with respect to the effects of observed covariates. For example \citet{ward2007disputes}, in a high profile example, argue that the the LSM improves inference about the effects of democracy, international trade, and participation in international organizations on the probability of inter-state conflict.

@@

%\section{Bayesian Inference and Priors in the Latent Space Model}

%Maximum likelihood estimation (MLE) is problematic in the LSM since the value of the likelihood function is invariant to rotation of the latent space \citep{hoff2002latent} (i.e., only the distances between points matter; not their absolute positions). Invariance to rotation means that the ML estimate is unidentified. As such, inference in the LSM is conducted using a Bayesian approach. Latent positions and other parameters are sampled using a Metropolis-Hastings algorithm. After sampling, latent space positions are rotated to match a common set of positions, via Procrustes transformation \citep{borg2005modern}. Procrustes transformation variance that is due to rotation of the latent space.  \cite{hoff2002latent} propose the use of diffuse independent normal priors for the latent positions and regression parameters. They also recognize the fact that isolates' (i.e., nodes with no ties) finite positions are only identified through the prior in the LSM. In one application they actually just exclude the isolates. Excluding isolates seems fine if the inferential purpose is to estimate positions in the latent space (for, e.g., visualization). However, if the goal is to conduct explanatory analysis that is representative of the full set of nodes, excluding isolates would amount to selecting on the dependent variable. Unlike in many practical applications of Bayesian inference, it is not possible to use an improper ``flat'' prior \citep{tibshirani1989noninformative} with the LSM, as an informative prior is necessary to assure finite latent positions.
%\begin{quote}
%We have not discussed in detail the choice of a prior distribution for latent positions in this article. Although simple, the diffuse independent normal priors presented in the examples may not accurately represent prior beliefs about the structure of social networks. More appropriate might be clustered point processes or mixtures of normals with an unknown number of components. Such priors could allow one to incorporate prior information on tendencies for clustering, without specifying cluster membership. This would add another level of hierarchy to the analysis, although the resulting model would be more flexible and perhaps more accurately represent any tendencies of populations to form segregating groups.
%\end{quote}
%Since it is necessary to use an informative prior with the LSM, we ask whether it is possible to specify the prior in a way that improves inference regarding covariate parameters. We propose to calibrate the prior to be on a scale comparable to the linear predictor estimated from a GLM. This scaling should discourage the latent space from replicating the linear predictor, encouraging the discovery of structure that remains in the data after identifying the effects of measured covariates.

%Our scaling rule is expressed as the prior variance for the latent positions, denoted $\nu^2$. We focus on a one-dimensional latent space $\mathbf{z}$ with a single dyadic covariate $\mathbf{x}$. The rule we derive generalizes naturally to more than one dimension, as the prior variance for each dimension could simply be set to $\nu^2/d$ where $d$ is the number of dimensions.  If we assume that the $\mathbf{z}$ are independent and distributed $\mathcal{N}(0,\sigma^2)$, it is straightforward to derive the distribution of $|z_i - z_j|$.\footnote{The assumption that $\mathbf{z} \sim \mathcal{N}(0,\sigma^2$) is consistent with $\mathbf{z}$ being drawn from the prior that is conventionally used in the LSM.} Since it is a special case of a ``normal difference distribution'', we know that $z_i - z_j \sim \mathcal{N}(0,2\sigma^2)$ \citep{devore2012}. The normality of $z_i - z_j$ implies that $|z_i - z_j|$ has a folded normal distribution \citep{leone1961}. Following from this result, we know that the variance of  $|z_i - z_j|$ is $\sigma^2(2-4/\pi)$.  Let $\theta^2$ be the empirical variance of $\mathbf{x}$. Then the variance in the linear predictor is $\beta^2\theta^2$. Given an estimate $\hat{\beta}$, we can set the variance of the latent position prior to $$\nu^2=\hat{\beta}^2\theta^2/(2-4/\pi).$$  This prior actively penalizes latent distance distributions that match or exceed the variance of the observed linear predictor. 
\subsection{Simulation Study}

@@

For simplicity and computational efficiency we consider a unidimensional latent space. We expect our results to generalize to higher dimensions and other variants of the LSM since each variant involves the use of unconstrained latent variables to represent unmeasured network structure. To generate an observed covariate that exhibits a controllable degree of colinearity with the latent network structure we follow a three-step process. First, we simulate unidimensional positions for each node, drawing from a normal distribution, and calculate the Euclidean distance $\mathbf{d}$ between each pair of positions. Second, given a target covariance matrix ($\Sigma$) among the covariates and distances, $\langle \mathbf{x}, \mathbf{d} \rangle$, we derive the conditional mean vector and covariance, assuming that $\mathbf{x}$ has a normal distribution given $\mathbf{d}$ (see \cite[pp. 116--117]{eaton1983} for the conditional normal derivation). Third, we simulate $x$ as a normal random variable with the respective conditional means and covariance. Finally we standardize $x$ to have zero mean and unit variance. To generate the covariance matrix $\Sigma$, which controls the dependence between the omitted network structure and the observed covariate $\mathbf{x}$ we utilize the C-vine method of \cite{lewandowski2009generating}. % fill in a basic description of the method here

>>>>

\noindent Due to the complex interdependence found in network data,
political networks scholars draw upon an increasingly sophisticated
toolkit for statistical inference. The latent space model (LSM) for
network data combines a generalized linear model with a latent spatial
embedding of the network. In numerous applications, researchers assume
that the latent spatial embedding can control for unmeasured
confounding structure that is related to the values of edges in the
network. There has been little to no research that considers the LSM's
performance in adjusting for unmeasured structure. We investigate the
LSM's performance via a simulation study. In the presence of an
unmeasured covariate that can be modeled using a latent space,
estimation and inferential error remain high under even moderate
confounding. However, the prediction error of the LSM when unmeasured
network structure is present is substantially lower in most cases. We
conclude that the LSM is most appropriate for exploration or
prediction.
%\footnote{This work was supported by National Science Foundation
%grants DGE-1144860, SES-1558661, SES-1619644, and CISE-1320219. Any
%opinions, findings, and conclusions or recommendations are those of
%the authors and do not necessarily reflect those of the sponsor.}
\end{abstract}
\thispagestyle{empty}
% \doublespacing
% Description of the possible challenges
\section{Introduction}

@@

The approach to adjusting for dependencies in the ERGM and SAOM is
quite similar to adjusting for confounding covariates in regression
modeling. The researcher specifies a set of dependencies that (s)he
hypothesizes to be important in the generative model for the
network. These dependencies are then explicitly included in a model
that simultaneously represents the effects of observed covariates
\citep{cranmer2011pa}. The LSM takes a different approach, which
involves the incorporation of latent variables to model network
structure. The LSM has an advantage over ERGM and SAOM in that
researchers need not develop a set of hypothesized dependencies in
order to model network structure that is not reflected in observed
covariates. However, this advantage hinges upon the capacity for the
LSM to discover unmodeled structure that could otherwise be attributed
to the observed covariates.

Despite their growing popularity, few studies exist that investigate
the performance of these models in adjusting for confounding network
structure. In the current study, we focus on the LSM, examining its
performance in reducing estimation and inferential errors regarding
the effects of observed covariates via adjustment for confounding
network structure. There are two reasons that using the LSM may lead
to increased error---one may result in Type 1 inferential error, and
the other in Type 2 error. First, if the unobserved (i.e., latent)
network structure is truly correlated with the observed covariates,
the unobserved structure that can be correlated with the observed
variable may be attributed to the observed variable. Under this
condition the latent space parameters are used to model sources of
variation that cannot be attributed to the observed covariate, and
inferences regarding the observed covariates remain subject to omitted
variable bias, which leads to an inflated Type 1 inferential error
rate. Second, the latent configurations inferred may result in a
representation of the network wherein a node's position in the latent
space is spuriously correlated with the observed covariates. Inferring
latent variables that are correlated with observed covariates would
lead to reduced efficiency in estimating observed covariate effects,
and result in a high Type 2 error rate.

Recent work on the LSM has explicitly addressed the potential problem
of Type 2 error. The AMEN framework for latent variable modeling of
networks is explicitly designed to model structure in the
residuals---structure that is leftover after accounting for the
observed covariates. \citet[p. 43]{hoff2015dyadic} notes that the
latent factors represent patterns in the network that, ``aren't
explained by the known regressors.''
\citet[pp. 12--13]{minhas2016inferential} also describe how AMEN is
designed such that the multiplicative effects, ``capture higher-order
dependence patterns that are left over [in the stochastic linear
  predictor] after accounting for any known covariate information.''
The approach incorporated in AMEN is effective at avoiding correlation
between the observed covariates and latent factors, which would result
in Type 2 error. However, there have been no methodological
innovations to avoid Type 1 error in latent variable modeling with
networks. Through a simulation study we find that the LSM, as
implemented in both the R \citep{R} packages \texttt{latentnet}
\citep{latentnet} and \texttt{amen} \citep{amen}, performs poorly in
terms of Type 1 error when the unmeasured network structure confounds
the relationship between the observed covariate and the network.

% CUT: not sure where to put this
%Despite the power of the LSM in representing network structure that is
%not modeled by the observed covariates, and its performance in
%predictive ties predicting ties, we have reason to doubt the LSM's
%capacity to adjust for confounding network structure.

\section{The Latent Space Model}

The LSM, introduced by \citet{hoff2002latent}, is used to estimate the
effect of covariates in the presence of latent network structure. Here
the distance function $|z_i - z_j|$ represents latent network
structure as homophily with respect to latent variables.  The distance
function is additively combined with a regression on observed dyadic
covariates, $x_{ij}$, to form a linear predictor for tie prediction.
As with a GLM, a link function, $g^{-1}$, maps this linear predictor
to the appropriate edge distribution, giving

@@

This ``Euclidean'' LSM is the original form proposed by
\citet{hoff2002latent}, and, as far as we can tell, the most commonly
used specification in the literature.

% CUT: what's the point of this paragraph?
%However, the LSM has been extended in several important
%directions. The latent space has been represented as a $k$-dimensional
%Euclildean space and by latent factors with a multiplicative metric
%\citep{hoff2002latent, hoff2004modeling, hoff_2005_jasa,
%  hoff2009multiplicative}. Additional structure has been introduced by
%adding random effects, which, for example, may involve sender or
%receiver specific effects for directed networks which capture
%differential activity rates amongst nodes
%\citep{hoff2003random}. Within this framework \cite{westveld2011mixed}
%models dynamic network data by treating the latent space as a
%stochastic process. \cite{handcock2007model} enable the LSM to model
%clustering that is not representable as homophily (i.e., stochastic
%equivalence) by combining the LSM with latent cluster
%models. \cite{hoff2008modeling} shows that the LSM and latent cluster
%models are special cases of an ``eigenmodel.'' That is, an
%eigendecomposition of a symmetric sociomatrix can be used to represent
%both latent space and cluster models, but not
%vice-versa. \citet{hoff2015dyadic} consolidates, extends, and
%implements several of these extensions in the \texttt{amen} package
%\citep{amen} for \R \citep{minhas2016inferential, hoff2015dyadic}.

@@

%Below we review several studies in which researchers have relied on
%the latent space modeling framework to adjust for unmeasured
%confounding structure.

Although the LSM has often been used as an exploratory or predictive
model, it has been applied in many cases for explanatory causal
modeling---to reduce estimation and/or inferential error with respect
to the effects of observed covariates. We do not claim that the
methodological literature in which the LSM was introduced or extended
has presented the LSM as capable of adjusting for unmeasured
confounding structure. However, as we document below, many researchers
who have applied the LSM claim that it has the capacity to adjust for
confounders, and use it for this purpose. For example
\citet{ward2007disputes}, in a high profile example, argue that the
the LSM improves inference about the effects of democracy,
international trade, and participation in international organizations
on the probability of inter-state conflict.

@@

\section{Simulation Study}

@@

For simplicity and computational efficiency we consider a
unidimensional latent space. We expect our results to generalize to
higher dimensions and other variants of the LSM since each variant
involves the use of unconstrained latent variables to represent
unmeasured network structure. To generate an observed covariate that
exhibits a controllable degree of colinearity with the latent network
structure we follow a three-step process. First, we simulate
unidimensional positions for each node, drawing from a normal
distribution, and calculate the Euclidean distance $\mathbf{d}$
between each pair of positions. Second, given a target covariance
matrix ($\Sigma$) among the covariates and distances, $\langle
\mathbf{x}, \mathbf{d} \rangle$, we derive the conditional mean vector
and covariance, assuming that $\mathbf{x}$ has a normal distribution
given $\mathbf{d}$ (see \cite[pp. 116--117]{eaton1983} for the
conditional normal derivation). Third, we simulate $x$ as a normal
random variable with the respective conditional means and
covariance. Finally we standardize $x$ to have zero mean and unit
variance. To generate the covariance matrix $\Sigma$, which controls
the dependence between the omitted network structure and the observed
covariate $\mathbf{x}$ we utilize the C-vine method of
\cite{lewandowski2009generating}.
